---
title: "Data Visualization Project"
author: "Vaneck Fokou"
date: "08/02/2021"
output: html_document
---

```{r setup, include=FALSE}
## install packages if necessary
list.of.packages <- c("janitor", "ggplot2", "dplyr", "lubridate", "zoo", "forcats", "ggthemes", "knitr", "rmarkdown")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)
library(directlabels)

## set working directory
knitr::opts_knit$set(root.dir = '/cloud/project/dataviz_project')
```

# Part 1: Exploratory Data Analysis

The data we'll be using for this part of the project were downloaded from [kaggle](https://www.kaggle.com/aaronschlegel/seattle-pet-licenses/data), and include information about "pet licenses issued by the Seattle Animal Shelter between 2005 and early 2017." We'll be exploring these data and generating a few exploratory plots in the first part of the project.


## The Data

First, we'll read the data in from our `data/raw_data` directory.

```{r pet-data, warning=FALSE} 
pets <- read.csv("data/raw_data/seattle_pet_licenses.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
```

## Explore the Data

```{r pet-explore, warning=FALSE} 
##  explore the data 
glimpse(pets)
```
```{r}
# number of unique animal's name

length(unique(pets$animal_s_name))
```


## Visualize the Data

```{r pet-barchart, warning=FALSE} 
## visual breakdown of how many of each species 
## are included in the dataset
a <- ggplot(pets) +
  geom_bar(aes(x=species, fill = species)) + 
  scale_fill_manual(values = c("magenta", "royalblue", "orange")) + 
  theme_bw()
a
```

```{r}
# save the figure
ggsave(plot = a, filename = "figures/exploratory_figures/species.png")
```


```{r pet-cat-table, warning=FALSE} 
## Table: Most frequent Cat Name 
pets %>% 
  filter(species == "Cat", animal_s_name != "") %>%
  group_by(animal_s_name) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Top 10 Cat Names in Seattle")
```

```{r pet-dog-table, warning=FALSE} 
## Table: Most frequent Dog Name
pets %>% 
  filter(species == "Dog", animal_s_name != "") %>%
  group_by(animal_s_name) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Top 10 Dog Names in Seattle")
```

```{r pet-linechart, warning=FALSE} 
## add date and ym columns
pets$date <- lubridate::ymd_hms(pets$license_issue_date)
pets$ym <- as.yearmon(pets$date, "%y%m")

## how the number of licenses recorded has changed over time
b <- pets %>%
  ## group by yearmonth (`ym`)
  group_by(ym) %>% 
  ## count number within each group
  summarise(n=n()) %>%
  ggplot(., aes(ym, n)) + 
  ## geom name for line chart
  geom_line(color = "red") +
  scale_x_yearmon() + 
  xlab("") + 
  ylab("Number of licenses")
b
```

```{r}
# saving the image 

ggsave(plot = b, filename = "figures/exploratory_figures/total_licences.png")
```


```{r pet-plot, warning=FALSE} 
### most frequent cat's primary breed 

pets %>%
  filter(species == "Cat", primary_breed != "") %>%
  group_by(primary_breed) %>%
  summarise(n = n()) %>%
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Most frequent Cat's primary breed")

### most frequent cat's secondary breed  
pets %>%
  filter(species == "Cat", secondary_breed != "") %>%
  group_by(secondary_breed) %>%
  summarise(n = n()) %>%
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Most frequent Cat's secondary breed")
```

```{r}
# Most frequent Dog's primary breed
pets %>%
  filter(species == "Dog", primary_breed != "") %>%
  group_by(primary_breed) %>%
  summarise(n = n()) %>%
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Most frequent dog's primary breed")

# Most frequent Dog's secondary breed
pets %>%
  filter(species == "Dog", secondary_breed != "") %>%
  group_by(secondary_breed) %>%
  summarise(n = n()) %>%
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Most frequent dog's secondary breed")
```


# Part 2: Explanatory Data Analysis

The data used in this part of the project were dowloaded from [FiveThirtyEight](https://data.fivethirtyeight.com/) - `steak-survey`.  They were originally used in the article: [How Americans Like Their Steak](https://fivethirtyeight.com/features/how-americans-like-their-steak/). The goal of this part of the project will be to recreate the data visualization used in this article.

## The Data

```{r meat-data, warning=FALSE} 
## read in the data
steak <- read.csv("data/raw_data/steak-risk-survey.csv", stringsAsFactors = FALSE) %>%
  ## remove first row which just includes the word "Response" in each column
  slice(2:n())
```

## Explore the Data

```{r meat-explore, warning=FALSE} 
## Explore the data 
glimpse(steak)
head(steak)
```

```{r}
# Number of people that eat steak

steak %>% group_by(Do.you.eat.steak.) %>%
  summarise(n = n())
```
```{r}
##
unique(steak$How.do.you.like.your.steak.prepared.)
```


## Wrangle the Data

```{r meat-wrangle, warning=FALSE} 
## wrangle the steak data set
pref <- steak %>% 
  mutate(steak_pref = factor(How.do.you.like.your.steak.prepared.,
                             levels = c("Well",
                                        "Medium Well",
                                        "Medium",
                                        "Medium rare",
                                        "Rare"))) %>% 
  filter(steak_pref != "") %>% 
  group_by(steak_pref) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))
```
```{r}

glimpse(pref)
```


## Visualize the Data

```{r meat-visualize, warning=FALSE}
## generate the plot 
p <- ggplot(pref) +
  ## specify you want to generate a bar chart
  geom_bar(aes(x = steak_pref, y = prop, fill = steak_pref),
                stat = 'identity', 
                width = 0.7) + 
  ## this adds text labels (you don't have to change anything here)
  geom_text(aes(label = paste0(as.integer(prop*100),"%"),
                x = steak_pref,
                y = prop), 
            stat= "identity", 
            hjust = -0.2, 
            size = 5,
            color = "grey40") +
  ## flip coordinates to make horizontal box plot
  coord_flip() +
  ## change the colors of the bars
  scale_fill_manual(values = c("#370B01", 
                           "#70402C",
                           "#955732", 
                           "#B36153",
                           "#B5413F")) +
  ## change the scale/labels of the steak-wellness variable (x-axis)
  scale_x_discrete(limits = levels(fct_rev(pref$steak_pref)),
                labels = c("Well", 
                           "Medium Well" = "Medium-well",
                           "Medium",
                           "Medium rare"="Medium-rare",
                           "Rare")) +
  ## change the scale/lables of the percent axis (y-axis)
  scale_y_discrete(labels = scales::percent, 
                expand = c(mult = c(0,0), 
                           add = c(0,0.035))) +
  ## change the title, subtitle, and caption
  labs(title="How Do You Like Your Steak Prepared?", 
                subtitle="From a survey of 432 steak-eating Americans",
                caption="FiveThirtyEight: Surveymonkey") +
  ## change the theme (use ggthemes)
  theme_economist() +
  ## fine tune the theme
  theme(axis.text = element_text(size = 14), 
                title = element_text(size = 16), 
                legend.position="none", 
                plot.caption=element_text(size = 12), 
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.text.x = element_blank(), 
                
  ) + xlab("") + ylab("")

p
```

## Save the Plot

```{r save-plot, warning=FALSE}
## save plot to figures/explanatory_figures directory
ggsave(plot = p, filename = "figures/explanatory_figures/steak_R.png", width = 8, height = 4)
```

## Visualize the Data

```{r}
head(pets)
```


```{r meat-plot, warning=FALSE}
## add date and ym columns
#pets$date <- lubridate::ymd_hms(pets$license_issue_date)
#pets$ym <- as.yearmon(pets$date, "%y%m")

## how the number of licenses of cats and dogs recorded has changed over time
g <- pets %>%
  ## group by yearmonth (`ym`)
  group_by(ym, species) %>% 
  ## count number within each group
  summarise(n=n()) %>%
  ggplot(., aes(ym, n, color = species)) + 
  ## geom name for line chart
  geom_line() + 
  scale_color_manual(values = c("orange", "brown", "red")) +
  scale_x_yearqtr(limits = c(2012, 2017),
                  format = "%Y", breaks=c(2012:2017), expand = c(0.1, 0.01)) + 
  labs(title = "Number of licenses of cats and dogs ", subtitle = "In the Seattle Animal Shelter between 2012 and 2017", caption = "Seattle Animal Shelter") +
  xlab("") + 
  ylab("Number of licenses") + 
  theme_minimal() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor.x = element_blank(), 
        legend.position = "none") +
  geom_dl(aes(label = species), 
          method = list(c("last.points"),
                        aes(colour = "black"), 
                        cex = 0.65))
g 
```

## Save this plot 

```{r save-meat-plot, warning=FALSE}
## save plot to figures/explanatory_figures directory
ggsave(plot = g, filename = "figures/explanatory_figures/pets_R.png")
```
